Burning sage: Reversing the curse of dimensionality in the visualization of high-dimensional data

### 2021-07-15 paper/paper.Rmd
Working on the revision

Paradoxically, the curse of dimensionality inverts for dimension reduction, resulting in an excessive amount of observations near the center of the distribution. This affects visualizations made on low-dimensional projections, like the tour [@As85;@BCAH05]. The effect is described by @diaconis1984, that most low-dimensional linear projections are approximately Gaussian, with observations concentrating in the center. This has motivated the development of indexes for projection pursuit which search for departure from normality. It is also related to what is called "data piling" in high-dimension low-sample size data [@10.2307/27639976;@10.1093/biomet/asp084]: all observations can collapse into a single point.  These issues also persist with non-linear dimension reduction techniques, and are often referred to as the "crowding problem", which methods like t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne] aim to alleviate.

Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections. Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions. This is a structural problem related purely to the geometry of projections from high-dimensional spaces. We show that it can be alleviated with a radial transformation on the low-dimensional space. Figure \ref{fig:density} illustrates this crowding problem, as would be experienced when using a tour. Two-dimensional linear projections of data, generated by sampling uniformly within $p$-dimensional hyperspheres ($p=3, 10, 100$), are displayed as hexbin plots. Color indicates log count of the bin, with yellow mapping the highest counts. As $p$ increases the density concentrates in the center of the projection. (Note that, sampling from a uniform in a sphere is used for this illustration for several reasons. Any linear projection is equivalent because of the rotation invariance of the distribution. The uniform sampling places focus on the space, points are spread evenly across the volume, and not a data distribution. Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the middle plot.)

@@

1. Obtain projection matrix $\mathbf{A}$. For $d$-dimensional projections this is an orthonormal $p \times d$ matrix. To ensure the smooth rotation of projections, each new $\mathbf{A}$ is obtained as an interpolated step in the sequence, as explained in @BCAH05.
2. Project the data by computing $\mathbf{Y} = \mathbf{X}\cdot\mathbf{A}$.
3. Map $\mathbf{Y}$ to the display to re-draw the projected data. For $d=2$ this typically maps the projected points onto a scatter plot display. With the new display we first transform $\mathbf{Y}$ as:
    + Center the 2-dimensional matrix $\mathbf{Y}$ and compute its polar coordinate representation $(r_y, \theta_y)$.
    + For each observation, first use Eq. \ref{eq:cutR} to get the trimmed radius $r_y^{\mathrm{trim}}$  within the specified range, and then apply the radial transformation defined in Eq. \ref{eq:resc} to obtain $r'_y$.
    + Use the transformed radial coordinate $r'_y$ to re-compute the mapping onto Euclidean coordinates $(y_1, y_2)$.
4. To fit the final projection onto the plotting canvas ranging between $[-1,1]$, we rescale each mapped observation $y$ using a scaling parameter $s$.

>>>>

Paradoxically, the curse of dimensionality inverts for dimension reduction, resulting in an excessive amount of observations near the center of the distribution. This affects visualizations made on low-dimensional projections. 
The effect is described by @diaconis1984, \textcolor{red}{showing that most linear projections onto low-dimensional subspaces are approximately Gaussian, and that observations concentrate near the center as the dimensionality of the full space increases.}
This has motivated the development of indexes for projection pursuit which search for departure from normality. It is also related to what is called "data piling" in high-dimension low-sample size data [@10.2307/27639976;@10.1093/biomet/asp084]: all observations can collapse into a single point.  These issues also persist with non-linear dimension reduction techniques, and are often referred to as the "crowding problem", which methods like t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne] aim to alleviate.

Here we are providing a solution to fix a long-standing problem in the tour [@As85;@BCAH05], where high-dimensional data is viewed by a movie of low-dimensional projections. Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions. This is a structural problem related purely to the geometry of projections from high-dimensional spaces. We show that it can be alleviated with a radial transformation on the low-dimensional space. Figure \ref{fig:density} illustrates this crowding problem, as would be experienced when using a tour. Two-dimensional linear projections of data, generated by sampling uniformly within $p$-dimensional hyperspheres ($p=3, 10, 100$), are displayed as hexbin plots. Color indicates log count of the bin, with yellow mapping the highest counts. As $p$ increases the density concentrates in the center of the projection. (Note that, sampling from a uniform in a sphere is used for this illustration for several reasons. Any linear projection is equivalent because of the rotation invariance of the distribution. The uniform sampling places focus on the space, points are spread evenly across the volume, and not a data distribution. Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the middle plot.)

@@

1. Obtain projection matrix $\mathbf{A}$. For $d$-dimensional projections this is an orthonormal $p \times d$ matrix. To ensure the smooth rotation of projections, each new $\mathbf{A}$ is obtained as an interpolated step in the sequence, as explained in @BCAH05.
2. Project the data by computing $\mathbf{Y} = \mathbf{X}\cdot\mathbf{A}$.
3. Map $\mathbf{Y}$ to the display to re-draw the projected data. For $d=2$ this typically maps the projected points onto a scatter plot display. With the new display we first transform $\mathbf{Y}$ as:
    + Center the 2-dimensional matrix $\mathbf{Y}$ and compute its polar coordinate representation $(r_y, \theta_y)$.
    + For each observation, first use Eq. \ref{eq:cutR} to get the trimmed radius $r_y^{\mathrm{trim}}$  within the specified range, and then apply the radial transformation defined in Eq. \ref{eq:resc} to obtain $r'_y$.
    + Use the transformed radial coordinate $r'_y$ to re-compute the mapping onto Euclidean coordinates $(y'_1, y'_2)$.
4. To fit the projected and transformed data $\mathbf{Y}'$ onto the plotting canvas ranging between $[-1,1]$, we rescale each mapped observation $y$ using a scaling parameter $s$.

### 2021-07-15 paper/paper.tex
Working on the revision

Paradoxically, the curse of dimensionality inverts for dimension
reduction, resulting in an excessive amount of observations near the
center of the distribution. This affects visualizations made on
low-dimensional projections, like the tour \citep{As85, BCAH05}. The
effect is described by \citet{diaconis1984}, that most low-dimensional
linear projections are approximately Gaussian, with observations
concentrating in the center. This has motivated the development of
indexes for projection pursuit which search for departure from
normality. It is also related to what is called ``data piling'' in
high-dimension low-sample size data
\citep{10.2307/27639976, 10.1093/biomet/asp084}: all observations can
collapse into a single point. These issues also persist with non-linear
dimension reduction techniques, and are often referred to as the
``crowding problem'', which methods like t-Distributed Stochastic
Neighbor Embedding (t-SNE) \citep{tsne} aim to alleviate.

Here we are providing a solution to fix a long-standing problem in the
tour, where high-dimensional data is viewed by a movie of
low-dimensional projections. Points crowding the center of the drawing
canvas detract from reading the plots, even with relatively few
dimensions. This is a structural problem related purely to the geometry
of projections from high-dimensional spaces. We show that it can be
alleviated with a radial transformation on the low-dimensional space.
Figure \ref{fig:density} illustrates this crowding problem, as would be
experienced when using a tour. Two-dimensional linear projections of
data, generated by sampling uniformly within \(p\)-dimensional
hyperspheres (\(p=3, 10, 100\)), are displayed as hexbin plots. Color
indicates log count of the bin, with yellow mapping the highest counts.
As \(p\) increases the density concentrates in the center of the
projection. (Note that, sampling from a uniform in a sphere is used for
this illustration for several reasons. Any linear projection is
equivalent because of the rotation invariance of the distribution. The
uniform sampling places focus on the space, points are spread evenly
across the volume, and not a data distribution. Also, \(p=10\) would
already be considered high-dimensional, as seen by the piling observed
in the middle plot.)

@@

\textcolor{red}{At this point a few characteristics of the transformation should be noted. First, as already mentioned, it is approximately linear near the center, and thus will not affect shapes in that part of the graph. In addition, since only the radial component is transformed, angles between the points will not be altered. (This can also be shown explicitly, ). Note however that shapes will be affected further away from the center (in particular towards the maximum range $R$). For example, a linear relation might appear curved because of the non-linear nature of the transformation in that region of the projected space. Thus, while the interpretation is straightforward for points projected in the central region, care must be taken with the interpretation of features observed in the outer region of the projected distribution.}

@@

  \begin{itemize}
  \tightlist
  \item
    Center the 2-dimensional matrix \(\mathbf{Y}\) and compute its polar
    coordinate representation \((r_y, \theta_y)\).
  \item
    For each observation, first use Eq. \ref{eq:cutR} to get the trimmed
    radius \(r_y^{\mathrm{trim}}\) within the specified range, and then
    apply the radial transformation defined in Eq. \ref{eq:resc} to
    obtain \(r'_y\).
  \item
    Use the transformed radial coordinate \(r'_y\) to re-compute the
    mapping onto Euclidean coordinates \((y_1, y_2)\).
  \end{itemize}
\item
  To fit the final projection onto the plotting canvas ranging between
  \([-1,1]\), we rescale each mapped observation \(y\) using a scaling
  parameter \(s\).
\end{enumerate}

@@

>>>>

Paradoxically, the curse of dimensionality inverts for dimension
reduction, resulting in an excessive amount of observations near the
center of the distribution. This affects visualizations made on
low-dimensional projections. The effect is described by
\citet{diaconis1984},
\textcolor{red}{showing that most linear projections onto low-dimensional subspaces are approximately Gaussian, and that observations concentrate near the center as the dimensionality of the full space increases.}
This has motivated the development of indexes for projection pursuit
which search for departure from normality. It is also related to what is
called ``data piling'' in high-dimension low-sample size data
\citep{10.2307/27639976, 10.1093/biomet/asp084}: all observations can
collapse into a single point. These issues also persist with non-linear
dimension reduction techniques, and are often referred to as the
``crowding problem'', which methods like t-Distributed Stochastic
Neighbor Embedding (t-SNE) \citep{tsne} aim to alleviate.

Here we are providing a solution to fix a long-standing problem in the
tour \citep{As85, BCAH05}, where high-dimensional data is viewed by a
movie of low-dimensional projections. Points crowding the center of the
drawing canvas detract from reading the plots, even with relatively few
dimensions. This is a structural problem related purely to the geometry
of projections from high-dimensional spaces. We show that it can be
alleviated with a radial transformation on the low-dimensional space.
Figure \ref{fig:density} illustrates this crowding problem, as would be
experienced when using a tour. Two-dimensional linear projections of
data, generated by sampling uniformly within \(p\)-dimensional
hyperspheres (\(p=3, 10, 100\)), are displayed as hexbin plots. Color
indicates log count of the bin, with yellow mapping the highest counts.
As \(p\) increases the density concentrates in the center of the
projection. (Note that, sampling from a uniform in a sphere is used for
this illustration for several reasons. Any linear projection is
equivalent because of the rotation invariance of the distribution. The
uniform sampling places focus on the space, points are spread evenly
across the volume, and not a data distribution. Also, \(p=10\) would
already be considered high-dimensional, as seen by the piling observed
in the middle plot.)

@@

\textcolor{red}{At this point a few characteristics of the transformation should be noted. First, as already mentioned, it is approximately linear near the center, and thus will not affect shapes in that part of the graph. In addition, since only the radial component is transformed, angles between the points will not be altered. (This can also be shown explicitly, from mathematical reformulation of the angles between two points after the transformation.) Note however that shapes will be affected further away from the center (in particular towards the maximum range $R$). For example, a linear relation might appear curved because of the non-linear nature of the transformation in that region of the projected space. Thus, while the interpretation is straightforward for points projected in the central region, care must be taken with the interpretation of features observed in the outer region of the projected distribution.}

@@

  \begin{itemize}
  \tightlist
  \item
    Center the 2-dimensional matrix \(\mathbf{Y}\) and compute its polar
    coordinate representation \((r_y, \theta_y)\).
  \item
    For each observation, first use Eq. \ref{eq:cutR} to get the trimmed
    radius \(r_y^{\mathrm{trim}}\) within the specified range, and then
    apply the radial transformation defined in Eq. \ref{eq:resc} to
    obtain \(r'_y\).
  \item
    Use the transformed radial coordinate \(r'_y\) to re-compute the
    mapping onto Euclidean coordinates \((y'_1, y'_2)\).
  \end{itemize}
\item
  To fit the projected and transformed data \(\mathbf{Y}'\) onto the
  plotting canvas ranging between \([-1,1]\), we rescale each mapped
  observation \(y\) using a scaling parameter \(s\).
\end{enumerate}

@@

\textcolor{red}{Note here that adjusting the resolution using the burning sage transformation will lead to an overall decrease in densitiy near the center. When studying the separation between groups it is important to keep in mind that the effect is the same within and between clusters. Thus, if points within a cluster remain grouped while separation between clusters becomes more apparant, this can be interpreted as a true finding. Conversely, in the case of a visual artifact we expect a general reduction of density that does not lead to better separation between the clusters.}
